Boosting Higgs Pair Production in the $b\bar{b}b\bar{b}$ Channel at the LHC with Multivariate Techniques

### 2015-07-14 paper/sec-mva.tex
work on the paper

\section{Multivariate analysis}
\label{sec:mva}

From the cut-based analysis,
combining the significance of the three categories,
we end up with a signal significance $S/\sqrt{B}\simeq 1.2(2.3)$
with all backgrounds (only QCD $4b$) taken into account,
which is not enough to be able to claim observation
of Higgs pair production
in this channel, and even less to extract the trilinear Higgs  coupling
$\lambda$.
%
We now process  the $n$-tuples of the signal and background
events that survive the cut-based
analysis by means of a multivariate
analysis, in order increase the signal over background
significance.
%
Multivariate techniques are by now a common tool to enhance signal
significance in HEP analysis, and this is fully justified
since they open a new window to improve the performance
of many physics analysis and searches.
%
For instance, the UCL $hh\to 4b$ analysis~\cite{Wardrope:2014kya}
uses a Boosted
Decision Tree MVA to improve signal discrimination.

In this section first of all we present the specific MVA that we use,
based on feed-forward multi-layer neural networks.
%
We then describe the input variable to the MVA, including the jet substructure
variables, and then present the results of $S/sqrt{B}$ including
the MVA effects.
%
Since we find that jet substructure variables are a crucial input
to the MVA, we end this section showing various kinematic distributions
of these variables.

{\bf add more refs from ATLAS and CMS}

@@

The classification of events as arising from either signal or
background processes via the use of a multivariate analysis is a common
technique in high energy physics~\cite{Baldi:2014pta,Wardrope:2014kya}.
%
In our work we use a specific type of  MVA to
disentangle signal and background events in the $hh\to 4b$ process,
in particular we use a multi-layer feed-forward artificial neural network (ANN),
also known as a {\it perceptron}.\footnote{This type of ANNs are the same
  as those used to parametrize Parton Distribution Functions
in the NNPDF global analyses~\cite{DelDebbio:2004qj,Ball:2008by,Ball:2011mu,Ball:2010de}.}
%
The basic idea is that the ANN is trained on an input composed on
all  signal and background
events which satisfy the
cut-based analysis, allowing to identify in an automated way
which are the most relevant variables to discriminate between the two.

@@

Let us now present the results of the MVA applied to the event samples
that satisfy the cut-based analysis discussed in the previous section.
%
First of all in Fig.~\ref{fig:nnresponse} we show the distribution of
the ANN output, after the training for signal and background events, for the three
exclusive categories.
%
All distributions have been normalized, so that their integral
  adds up to one.
%
Given that the final separation between signal and background will be achieving by introducing
a cut in the ANN output, the more separated that the two distributions are, the more efficient
the MVA discrimination will be.

@@

From Fig.~\ref{fig:nnresponse} we see that in the boosted category the MVA manages
to achieve a clear discrimination between signal and background, with the two distributions
nicely peaking at 1 and at 0 respectively.
%
This indicates that introducing a suitable cut in the ANN output will substantially reduce the background,
while keeping a reasonable signal efficiency.
%
The performance of the MVA discrimination is similar though slightly worse in the intermediate
category, and it is markedly worse in the resolved category.
%
We have traced back these differences to the use of jet substructure variables
in the boosted and intermediate categories, which induce the maximum
discrimination between signal and background events.

The results for the signal selection efficiency and the 
background rejection rate as a function of the cut in the ANN output
define the so-called  Receiver-Operating Characteristic (ROC)
curve.
%
It is clear that we can achieve a very high signal efficiency by using
a small value of the cut in the ANN output, but this choice will be
affected from a poor background
rejection.
%
Conversely, using a higher value of the cut will increase background rejection at the
cost of dropping signal efficiency.
%
The usefulness of ROC curves is that they allow to determine the
optimal cut value that removes enough background without degrading too much the signal efficiency.
%
In addition, it is important to select a value of the cut in ANN output so that
the number of expected signal events is still large enough.

@@

%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
\begin{figure}[t]
\begin{center}
\includegraphics[width=0.65\textwidth]{plots/roc.pdf}
\caption{\small ROC curve for the background rejection rate as a function of the signal
  selection efficiency as the cut in the ANN output is varied.
  %
  We show the results for the three exclusive categories: boosted, intermediate
  and resolved.
}
\label{fig:exampleroc}
\end{center}
\end{figure}
%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%

From Fig.~\ref{fig:exampleroc} we see that, always starting from the signal
and background events that survive the basic cut-based analysis,
one can achieve a  signal
efficiency of $\sim 80\%$ while reducing the background by
$60\%$, $55\%$ and $45\%$ in the boosted, intermediate and resolved
categories respectively.
%
It is easy to translate these factors into an  improvement factor of the pre-MVA
signal significance $S/\sqrt{B}$.
%
For example, requiring a signal efficiency of 90\%, we find a background rejection
rate of $40\%$, $30\%$ and $20\%$ respectively, which corresponds to an increase
in $S/\sqrt{B}$ of 1.2, 1.1 and 1.0 respectively.

Of course, and important restriction on the range of allowed cuts in the ANN output
is that the cut cannot be so hard that the actual number of signal events expected
at the HL-LHC becomes unreasonably small.
%
To verify how many signal and background events are left after the MVA cut,
we show in Fig.~\ref{fig:nev2} the number of signal (dashed lines) and background (solid lines)
  events expected at the HL-LHC as a function of the cut in the ANN output,
  for the three separate categories.
  %
  Note that the value of the cut in the ANN cut can be separately optimised in the three
  categories, even if here we plot them together for easiness of presentation.

@@

As can be seen from Fig.~\ref{fig:nev2}, in the boosted category
event with a cut in the ANN output as hard as 0.9 we are still left
with around 100 events at the HL-LHC, which is a reasonable number,
which the background being reduced substantially to around
500 events.
%
A similar cut would work fine for the intermediate category, though here
the backgrounds are much larger and thus the expected signal significance smaller.
%
For the resolved case, the signal significance would be only slightly
improved for any value of the cut.

A very useful property of MVA's such as the one used in this work
is that they provide direct  physical insight about which of the
input variables contribute the most to the signal over background
separation.
%
In the case of ANNs, this can be quantified by computing the sum
of the absolute value of al the weights connected to a given
input neuron $i$, that is
\be
\label{eq:totweight}
\omega^{\rm tot}_i \equiv \sum_{k=1}^{n^{(2)}} \Big|\omega^{(2)}_{ki}\Big| \, ,
\qquad i=1,\ldots,N_{\rm var} \, ,
\ee
with $\omega^{(2)}_{ki}$ the value of the weight connecting
the $k$-th neutron of the second layer with the $i$-th neuron of
the first (input) layer, and $n^{(2)}=5$ the number of
neurons in the second layer.
%
The input variables with a larger value of $\omega^{\rm tot}_i$ will those
be the ones that play a dominant role in enhancing the signal
discrimination using the MVA.

@@

\subsection{Another look at the jet substructure variables}

It is possible to verify that the usefulness of the MVA when
applied to the boosted category arises from the 
the information contained on the various jet substructure variables.
%
From the distribution of total associated ANN weights in the
boosted category, shown in Fig.~\ref{fig:nnweights}, we see that
all substructure variables contain useful information:
the $k_t$ splitting scales $\sqrt{d_{12}}$,
the $\tau_{21}$ subjettiness ratio, and the ECF ratios
$C_2^{{\beta}}$ and $D_2^{{\beta}}$.
%
Interestingly, for some substructure variables, such as $\tau_{21}$,
the total associated weight turns out to be stronger in the leading
than in the sub-leading Higgs candidate (or vice-versa), indicating some
degree of redundancy when the same variable is used for the two Higgs candidates
at the same time.

>>>>

\section{Results from the multivariate analysis}
\label{sec:mva}

From the cut-based analysis,
combining the significance of the three categories,
we end up with a signal significance of $S/\sqrt{B}\simeq 1.2(2.3)$
with all backgrounds (only QCD $4b$) taken into account.
%
This does not seem
to be enough to be able to claim observation
of Higgs pair production
in this channel, and even less to extract the trilinear Higgs  coupling
$\lambda$ with good precision.
%
In this section we show how the signal significance for
our process can
be substantially improved when
applying a multivariate analysis (MVA) to the events
 that survive the cut-based
 analysis.

%
Multivariate techniques are by now a common tool to enhance signal
significance in HEP analysis, and this is fully justified
since they open a new window to improve the performance
of many physics analysis and searches.
%
In particular, the classification of events as arising from either signal or
background processes via the use of a multivariate analysis is becoming
a mature
technique in LHC
applications~\cite{Baldi:2014pta,Aaltonen:2012qt,
  Wardrope:2014kya,Chatrchyan:2013zna,Dall'Osso:2015aia}.
%
For instance, the CMS $Vh(\to b\bar{b})$ analysis~\cite{Chatrchyan:2013zna}
is based on a Boosted Decision Tree (BDT) MVA, and the UCL $hh\to 4b$
feasibility study~\cite{Wardrope:2014kya}
also used  a BDT to improve signal discrimination.

In this section, first of all we present the specific MVA that we use,
based on feed-forward multi-layer neural networks.
%
We then introduce the input variables that are
used the MVA, including the jet substructure
variables, and then present the results of $S/\sqrt{B}$ including
the MVA effects.
%
Since we find that jet substructure variables are a crucial input
to the MVA, we end this section showing various kinematic distributions
of these variables.

@@

%
In our work we use a specific type of  MVA to
disentangle signal and background events in the $hh\to 4b$ process,
a multi-layer feed-forward artificial neural network (ANN),
also known as a {\it perceptron}.\footnote{This type of ANNs are the same
  as those used to parametrize Parton Distribution Functions
in the NNPDF global analyses~\cite{DelDebbio:2004qj,Ball:2008by,Ball:2011mu,Ball:2010de}.}
%
This family of ANNs is sometimes known as {\it deep} neural networks.
%
The basic idea is that the ANN is trained on an input composed on
all  signal and background
events which satisfy the requirements of the
cut-based analysis.
%
The main result of the trained ANN is identifying, in a fully automated way,
which are the most relevant variables to discriminate between the two.

@@

In Fig.~\ref{fig:nnresponse} we show the distribution of
the ANN output at the end of the training, for the exclusive
boosted, intermediate and resolved categories.
%
The distributions are normalized so that their integral
  adds up to one.
%
The  separation between signal and background is achieving by introducing
a cut, $y_{\rm cut}$, in the ANN output, so that MC events with $y_i\ge
y_{\rm cut}$ are classified as signal events, and those with
 $y_i <
y_{\rm cut}$ as background events.
%
Therefore,
the more separated that the two distributions are, the more efficient
the MVA discrimination will be.

@@

From Fig.~\ref{fig:nnresponse} we see that in the boosted category the MVA manages
to achieve a clear discrimination between signal and background, with the two distributions
nicely peaking at 1 and at 0 respectively.
%
This indicates that introducing a suitable cut
$y_{\rm cut}$
in the ANN output will substantially reduce the background,
while keeping a reasonable signal efficiency.
%
The performance of the MVA discrimination is similar though slightly worse in the intermediate
category, and it is markedly worse in the resolved category.
%
We have traced back these differences to the use of jet substructure variables
in the boosted and intermediate categories, whose use substantially
improves the
discrimination between signal and background events.

The results for the signal selection efficiency and the 
background rejection rate as a function of the cut in the ANN output
$y_{\rm cut}$
define the so-called  Receiver-Operating Characteristic (ROC)
curve.
%
It is clear that we can achieve  high signal efficiency by using
a small value of the cut in the ANN output, but this choice will be
affected from a poor background
rejection.
%
Conversely, using a higher value of the cut will increase background rejection at the
cost of dropping signal efficiency.
%
The usefulness of ROC curves is that they allow to determine and
optimal cut value that removes enough background without degrading too much the signal efficiency.
%
Another important requirement on the value of $y_{\rm cut}$ is that
the  number of expected  signal events (the physical
events for a given integrated
luminosity, not the MC events used for the ANN training)
after the cut is large enough.

@@

%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
\begin{figure}[t]
\begin{center}
\includegraphics[width=0.65\textwidth]{plots/roc.pdf}
\caption{\small ROC curve for the background rejection rate as a function of the signal
  selection efficiency, as the cut in the ANN output is varied.
  %
  We show the results for the three exclusive categories: boosted, intermediate
  and resolved.
}
\label{fig:exampleroc}
\end{center}
\end{figure}
%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%

From Fig.~\ref{fig:exampleroc} we see that, always starting from the signal
and background events that survive the basic cut-based analysis,
one can achieve a  signal
efficiency of $\sim 90\%$ while reducing the background by
$80\%$, $60\%$ and $50\%$ in the boosted, intermediate and resolved
categories respectively.
%
It is easy to translate these factors into an  improvement factor of the pre-MVA
signal significance $S/\sqrt{B}$.
%
As an illustration, if a signal efficiency of 90\%
is required, we find that the MVA improves $S/\sqrt{B}$
by  approximately 2.0, 1.4 and 1.3  in the three categories respectively.

As mentioned above, an
 important restriction on the range of allowed cuts in the ANN output
 is that $y_{\rm cut}$
 cannot be so hard that the actual number of signal events expected
at the HL-LHC becomes unreasonably small.
%
To verify how many signal and background events are left after the MVA cut,
we show in Fig.~\ref{fig:nev2} the number of signal (dashed lines) and background (solid lines)
  events expected at the HL-LHC as a function of $y_{\rm cut}$,
  for the three exclusive categories.
  %
  Note that the value of the cut in the ANN cut is separately optimised in the three
  categories.

@@

As can be seen from Fig.~\ref{fig:nev2}, in the boosted category
event with a cut in the ANN output as hard as $y_{\rm cut}\simeq 0.9$
we are still left
with around 100 events at the HL-LHC, which is a reasonable number,
which the background being reduced substantially to around
500 events.
%
A similar cut would work fine for the intermediate category, though here
the backgrounds are much larger and thus the expected signal significance smaller.
%
For the resolved case, the signal significance would be only slightly
improved for any value of the cut.

A very useful property of MVAs such as the one used in this work
is that they provide direct  physical insight about which of the
input variables contribute the most to the signal over background
separation.
%
In the case of ANNs, this can be quantified by computing the sum
of the absolute value of al the weights connected to a given
input neuron $i$, that is
\be
\label{eq:totweight}
\omega^{\rm tot}_i \equiv \sum_{k=1}^{n^{(2)}} \Big|\omega^{(2)}_{ki}\Big| \, ,
\qquad i=1,\ldots,N_{\rm var} \, ,
\ee
with $\omega^{(2)}_{ki}$ the value of the weight connecting
the $k$-th neutron of the second layer with the $i$-th neuron of
the first (input) layer, and $n^{(2)}=5$ the number of
neurons in the second layer.
%
The input variables with a larger value of $\omega^{\rm tot}_i$ will those
be the ones that play a dominant role in enhancing the signal
discrimination using the MVA.

@@

\subsection{Another look at jet substructure variables}

It is possible to verify that the usefulness of the MVA when
applied to the boosted category arises from the 
the information contained on the various jet substructure variables.
%
From the distribution of total associated ANN weights in the
boosted category, shown in Fig.~\ref{fig:nnweights}, we see that
all substructure variables contain useful information:
the $k_t$ splitting scales $\sqrt{d_{12}}$,
the $\tau_{21}$ subjettiness ratio, and the ECF ratios
$C_2^{(\beta)}$ and $D_2^{(\beta)}$.
%
Interestingly, for some substructure variables, such as $\tau_{21}$,
the total associated weight turns out to be stronger in the leading
than in the sub-leading Higgs candidate (or vice-versa), indicating some
degree of redundancy when the same variable is used for the two Higgs candidates
at the same time.
